### 2025-05-10 sec/3_tls_opt.tex
Moving TLS relax

\section{Wheighted Least Squares convex relaxation}

In this section we propose a novel convex relaxation for the Truncated Least Squares problem that can be computed in linear time.

@@

where $\mathbf{x}_i$ are data points and $r(\cdot)^2$ is some convex residual function such as the squared Euclidean distance $||\mathbf{x} - \mathbf{x}_i||_2^2$. In the following, we will refer to the i-th residual term $r(\mathbf{x}, \mathbf{x_i})^2$ simply as $r_i$.
The idea of the WLS-relaxation is the observation that this function is only non-convex because of the truncation -- the residuals are convex (parabolas). Since sums of convex functions are convex, we can focus on relaxing each term separately.

For this, we first find the minimum and maximum values of each residual $r_i$ is over the given interval, denoted as $r_{min}^i$ and $r_{max}^i$ respectively. The key geometric idea of our relaxation is to reweight these residuals such that after reweighting, their maximum becomes $\epsilon^2$ and thus they are not affected by truncation (Fig. \ref{fig:wlsrelaxideawithrelax}).

@@

In the following we show how $r_{min}$ and $r_{max}$ can easily be computed analytically for different models (i.e. translation and rotation), given spherical intervals, i.e. balls.

@@

This problem can be solved easily in closed form.
We first form the Lagrangian to account for the constraint:
\begin{equation}
	\begin{aligned}	
		L(\mathbf{x}, \lambda) =  \sum_{i=1}^{N} w_i ||\mathbf{d}_i - \mathbf{x}||_2^2 + \lambda (
		\mathbf{x}^\transposed \mathbf{x} - n_r^2)\\
	\end{aligned}
\end{equation}

then, we set the first partial derivatives to zero: 
\begin{equation}
	\begin{aligned}	
		\frac{\partial L}{\partial \mathbf{x}} &= \frac{\partial }{\partial \mathbf{x}} \left( \sum_{i=1}^{N} w_i \left(\mathbf{d}_i - \mathbf{x}\right)^\transposed \left(\mathbf{d}_i - \mathbf{x}\right) + \lambda (
		\mathbf{x}^\transposed \mathbf{x} - n_r^2) \right)\\
		&= - 2\sum_{i=1}^{N} w_i \left(\mathbf{d}_i - \mathbf{x}\right) + 2 \lambda \mathbf{x}\\
		&= \mathbf{0}\\
	\end{aligned}
\end{equation}

By rearranging, we can obtain therefore the solution in closed form:
\begin{equation}
	\label{eq:trans-qcqp-lagrangian}
	\begin{aligned}	
		-2  &\sum_{i=1}^{N}  w_i \left(\mathbf{d}_i - \mathbf{x}\right) + 2 \lambda \mathbf{x} = \mathbf{0}\\
		\iff \lambda \mathbf{x} &= \sum_{i=1}^{N} w_i \left(\mathbf{d}_i - \mathbf{x}\right)\\
		\lambda \mathbf{x} &= \sum_{i=1}^{N}  w_i \mathbf{d}_i  - \sum_{i=1}^{N}  w_i \mathbf{x}\\
		\lambda \mathbf{x} &= \sum_{i=1}^{N}  w_i \mathbf{d}_i  - \mathbf{x} \sum_{i=1}^{N}  w_i\\
		\left( \lambda + \sum_{i=1}^{N}  w_i \right) \mathbf{x} &= \sum_{i=1}^{N} w_i \mathbf{d}_i \\
		\mathbf{x} &= \frac{1}{\lambda + \sum_{i=1}^{N}  w_i} \sum_{i=1}^{N} w_i \mathbf{d}_i 
	\end{aligned}
\end{equation}

Now the Karush-Kuhn-Tucker (KKT) conditions require $\lambda \geq 0$ and $\lambda (
\mathbf{x}^\transposed \mathbf{x} - n_r^2) = 0$ in addition to satisfying the constraint, i.e. $|| \mathbf{x}||_2 \leq n_r$. It follows that in practice we obtain the optimum by first computing $\mathbf{x}$ with Eq. \ref{eq:trans-qcqp-lagrangian}, assuming $\lambda = 0$. Then we check whether the inequality constraint is satisfied, i.e. whether $|| \mathbf{x}||_2 \leq n_r$ holds, and if not, we simply normalise $\mathbf{x}$ so that $|| \mathbf{x}||_2 = n_r$, yielding the optimal solution. Intuitively, from Eq. \ref{eq:trans-qcqp-lagrangian} we already see that the direction of the vector does not change regardless of the optimal Lagrange multiplier $\lambda$ since it is only a scalar factor.

>>>>

\section{WLS Relaxation: A simple linear time relaxation}

In this section we propose a novel convex relaxation for the Truncated Least Squares problem that can be computed in linear time and can be minimized by solving a simple weighted least squares problem.

@@

where $\mathbf{x}_i$ are data points and $r(\cdot)^2$ is some convex residual function such as the squared Euclidean distance $||\mathbf{x} - \mathbf{x}_i||_2^2$. In the following, we will refer to the i-th residual term $r(\mathbf{x}, \mathbf{x_i})^2$ simply as $r_i$.
The idea of the WLS-relaxation is the observation that this function is only non-convex because of the truncation -- the residuals itself are convex. Since sums of convex functions are convex, we can focus on relaxing each term separately.

For this, we first find the minimum and maximum values of each residual $r_i$ is over a given interval, denoted as $r_{min}^i$ and $r_{max}^i$ respectively. We then re-weight these residuals such that their maximum becomes $\epsilon^2$ and thus they are not truncated over the entire interval (Fig. \ref{fig:wlsrelaxideawithrelax}).

@@

In the following we show how $r_{min}$ and $r_{max}$ can easily be computed analytically for different models (i.e. translation and rotation), given spherical intervals (balls).

@@

This problem can be solved easily in closed form, obtaining the optimal solution $\mathbf{x}^*$ with:

\begin{equation}
	\label{eq:trans-qcqp-sol}
	\begin{aligned}	
		\mathbf{x}^* &= \frac{1}{\lambda^* + \sum_{i=1}^{N}  w_i} \sum_{i=1}^{N} w_i \mathbf{d}_i 
	\end{aligned}
\end{equation}

where the optimal Lagrange multiplier $\lambda^*$ is chosen so that the norm of $\mathbf{x}^*$ is no greater than $n_r$ (see App. for proof).
It follows that in practice we obtain the optimum by first computing $\mathbf{x}$ with Eq. \ref{eq:trans-qcqp-lagrangian}, assuming $\lambda = 0$. Then we check whether the inequality constraint is satisfied, i.e. whether $|| \mathbf{x}||_2 \leq n_r$ holds, and if not, we simply normalise $\mathbf{x}$ so that $|| \mathbf{x}||_2 = n_r$, yielding the optimal solution. Intuitively, from Eq. \ref{eq:trans-qcqp-lagrangian} we already see that the direction of the vector does not change regardless of the optimal Lagrange multiplier $\lambda$ since it is only a scalar factor.

### 2025-05-10 sec/X_suppl.tex
Some polishing
Moving TLS relax

	\begin{equation}
		\label{eq:q-axis-angle}
		\begin{aligned}
			\mathbf{q} = \left( \sin \left( \frac{\theta}{2} \right) \mathbf{n}, \cos \left(\frac{\theta}{2} \right) \right)^\transposed
		\end{aligned}
	\end{equation}
	where $\mathbf{n}$ is a unit-vector that is the axis of rotation and $\theta$ is the angle of rotation.
	We rewrite the objective function $\mathbf{q}^\transposed \mathbf{K} \mathbf{q}$  of \ref{eq:davenports-q-problem} using the blocks of the block-matrix $\mathbf{K}$:

@@

>>>>

	\begin{equation}
		\label{eq:q-axis-angle}
		\begin{aligned}
			\mathbf{q} = \left( \sin \left( \frac{\theta}{2} \right) \mathbf{n}, \cos \left(\frac{\theta}{2} \right) \right)^\transposed
		\end{aligned}
	\end{equation}
	where $\mathbf{n} \in \Rthree$ is a unit-vector that is the axis of rotation and $\theta$ is the angle of rotation.
	We rewrite the objective function $\mathbf{q}^\transposed \mathbf{K} \mathbf{q}$  of \ref{eq:davenports-q-problem} using the blocks of the block-matrix $\mathbf{K}$:

@@

\begin{proof}
	We first form the Lagrangian to account for the constraint:
	\begin{equation}
		\begin{aligned}	
			L(\mathbf{x}, \lambda) =  \sum_{i=1}^{N} w_i ||\mathbf{d}_i - \mathbf{x}||_2^2 + \lambda (
			\mathbf{x}^\transposed \mathbf{x} - n_r^2)\\
		\end{aligned}
	\end{equation}

	then, we set the first partial derivatives to zero: 
	\begin{equation}
		\begin{aligned}	
			\frac{\partial L}{\partial \mathbf{x}} &= \frac{\partial }{\partial \mathbf{x}} \left( \sum_{i=1}^{N} w_i \left(\mathbf{d}_i - \mathbf{x}\right)^\transposed \left(\mathbf{d}_i - \mathbf{x}\right) + \lambda (
			\mathbf{x}^\transposed \mathbf{x} - n_r^2) \right)\\
			&= - 2\sum_{i=1}^{N} w_i \left(\mathbf{d}_i - \mathbf{x}\right) + 2 \lambda \mathbf{x}\\
			&= \mathbf{0}\\
		\end{aligned}
	\end{equation}

	By rearranging, we can obtain therefore the solution in closed form:
	\begin{equation}
		\label{eq:trans-qcqp-lagrangian}
		\begin{aligned}	
			-2  &\sum_{i=1}^{N}  w_i \left(\mathbf{d}_i - \mathbf{x}\right) + 2 \lambda \mathbf{x} = \mathbf{0}\\
			\iff \lambda \mathbf{x} &= \sum_{i=1}^{N} w_i \left(\mathbf{d}_i - \mathbf{x}\right)\\
			\lambda \mathbf{x} &= \sum_{i=1}^{N}  w_i \mathbf{d}_i  - \sum_{i=1}^{N}  w_i \mathbf{x}\\
			\lambda \mathbf{x} &= \sum_{i=1}^{N}  w_i \mathbf{d}_i  - \mathbf{x} \sum_{i=1}^{N}  w_i\\
			\left( \lambda + \sum_{i=1}^{N}  w_i \right) \mathbf{x} &= \sum_{i=1}^{N} w_i \mathbf{d}_i \\
			\mathbf{x} &= \frac{1}{\lambda + \sum_{i=1}^{N}  w_i} \sum_{i=1}^{N} w_i \mathbf{d}_i 
		\end{aligned}
	\end{equation}

	Now the Karush-Kuhn-Tucker (KKT) conditions require $\lambda \geq 0$ and $\lambda (
	\mathbf{x}^\transposed \mathbf{x} - n_r^2) = 0$ in addition to satisfying the constraint, i.e. $|| \mathbf{x}||_2 \leq n_r$.
\end{proof}
